---
title: "Hands-on Exercise 4d: Funnel Plots for Fair Comparisons"
author: "Sindy"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## 1. Overview

Funnel plot is a specially designed data visualisation for conducting unbiased comparison between outlets, stores or business entities.

### 1.1 Getting started

In this exercise, four R packages will be used.
```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

### 1.2 Importing Data

In this section, COVID-19_DKI_Jakarta will be used. The data was downloaded from Open Data Covid-19 Provinsi DKI Jakarta portal. For this hands-on exercise, we are going to compare the cumulative COVID-19 cases and death by sub-district (i.e. kelurahan) as at 31st July 2021, DKI Jakarta.

The code chunk below imports the data into R and save it into a tibble data frame object called covid19.

```{r}
covid19 <- read_csv("data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)
```





## 4. Reference

Credits to [Prof Kam](https://r4va.netlify.app/chap12).
