---
title: "Hands-on Exercise 2: Beyond ggplot2 Fundamentals"
author: "Sindy"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## 1. Getting started

### 1.1 Installing and loading the required libraries

In this exercise, beside tidyverse, four R packages will be used. They are:

-   ggrepel: an R package provides geoms for ggplot2 to repel overlapping text labels.
-   ggthemes: an R package provides some extra themes, geoms, and scales for 'ggplot2'.
-   hrbrthemes: an R package provides typography-centric themes and theme components for ggplot2.
-   patchwork: an R package for preparing composite figure created using ggplot2.

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 
```


### 1.2 Importing data

For the purpose of this exercise, a data file called *Exam_data* will be used. It consists of year end examination grades of a cohort of primary 3 students from a local school. It is in csv file format.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## 2. Beyond ggplot2 Annotation: ggrepel

One of the challenge in plotting statistical graph is annotation, especially with large number of data points.

::: panel-tabset
## Default

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## ggrepel

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID),
                   hjust = .5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

We simply replace `geom_label()` by `geom_label_repel()`.


## 3. Beyond ggplot2 Themes

ggplot2 comes with eight [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html), they are: `theme_gray()`, `theme_bw()`, `theme_classic()`, `theme_dark()`, `theme_light()`, `theme_linedraw()`, `theme_minimal()`, and `theme_void()`.

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=25, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +  
  ggtitle("Distribution of Maths scores") 
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=25, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```
:::

### 3.1 Working with ggtheme package

**ggthemes** provides ‘ggplot2’ themes that replicate the look of plots by Edward Tufte, Stephen Few, Fivethirtyeight, The Economist, ‘Stata’, ‘Excel’, and The Wall Street Journal, among others.

In the example below, The Economist theme is used.

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=25, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```


It also provides some extra geoms and scales for ‘ggplot2’.

::: panel-tabset

## Scatter plot
```{r}
ggplot(data = exam_data, aes(x = MATHS, y = ENGLISH, color = GENDER)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_economist() +
  ggtitle("English vs. Maths Scores by Gender") +
  theme_economist()
```
## Bar chart
```{r}
ggplot(data = exam_data, aes(x = MATHS, fill = GENDER)) +
  geom_bar(position = "dodge") +
  scale_fill_wsj() +
  ggtitle("Student Count by Maths and Gender") +
  theme_wsj() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```
:::

### 3.2 Working with hrbthems package

`hrbrthemes` package provides a base theme that focuses on typographic elements, including where various labels are placed as well as the fonts that are used.

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=25, 
                 color="grey25", 
                 fill="grey90") +
  scale_x_continuous(breaks = seq(0, 100, by = 25)) + # Set x-axis breaks at 25-unit intervals
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```
The second goal centers around productivity for a production workflow. In fact, this “production workflow” is the context for where the elements of hrbrthemes should be used. 

code here...



## 10. Reference

Credits to [Prof Kam](https://r4va.netlify.app/chap01#learning-outcome).
