---
title: "In-class Exercise 7: Visualising, Analysing and Forecasting Time-series Data: tidyverts methods"
author: "Sindy"
date: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## 1. Getting Started

### 1.1 Load Required R Packages

For the purpose of this in-class exercise, the following R packages will be used.

```{r}
pacman::p_load(tidyverse, tsibble, feasts, fable, seasonal)
```

### 1.2 Importing the data
```{r}
ts_data <- read_csv(
  "data/visitor_arrivals_by_air.csv")
```

In the code chunk below, dmy() of lubridate package is used to convert data type of Month-Year field from Character to Date.

```{r}
ts_data$`Month-Year` <- dmy(
  ts_data$`Month-Year`)
```

> Tip: make sure the format is correct (chr vs Date fields)

19.2.3 Conventional base ts object versus tibble object

```{r}
ts_data_ts <- ts(ts_data)       
head(ts_data_ts)
```
Different dataframes... (explanation here)
not a typical tibble dataframe
> class(ts_data_ts)
[1] "mts"    "ts"     "matrix" "array" 
> class(ts_data)
[1] "spec_tbl_df" "tbl_df"      "tbl"         "data.frame"

Use ts_data for data prep and conversion
then convert to ts_data_ts for a timeseries object?



some explanation here!
```{r}
ts_tsibble <- ts_data %>%
  mutate(Month = yearmonth(`Month-Year`)) %>%
  as_tsibble(index = `Month`)
```

> class(ts_tsibble)
[1] "tbl_ts"     "tbl_df"     "tbl"        "data.frame"

sth that can be used by dpylr and tidyr and also timeseries time series manner?



visual time series.,,,

transform the series for you to work with first
cause cannot see the header (numbers correspond to which country) - so transform to a long table. Prepare data by reordering data differently.
```{r}
ts_longer <- ts_data %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```


Visualising single time-series: ggplot2 methods

```{r}
ts_longer %>%
  filter(Country == "Vietnam") %>%
  ggplot(aes(x = `Month-Year`, 
             y = Arrivals))+
  geom_line(size = 0.5)
```

### plot multiple countries

```{r}
#| fig-height: 8
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals,
           color = Country))+
  geom_line(size = 0.5) +
  theme(legend.position = "bottom", 
        legend.box.spacing = unit(0.5, "cm"))
```


or use facet
you can define no of col --> rows will be automatically calculated...

```{r}
#| fig-height: 12
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals))+
  geom_line(size = 0.5) +
  facet_wrap(~ Country,
             ncol = 3,
             scales = "free_y") +
  theme_bw()
```

NOTE!!! intervals are not constant!!


```{r}
tsibble_longer <- ts_tsibble %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```

Useful to start with
```{r}
#| fig-height: 6
tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  autoplot(Arrivals) + 
  facet_grid(Country ~ ., scales = "free_y")
```


shows you the distribution.
e.g. Italians only come in at a big crowd in the month of Aug. Rest of the year hover around 5000.

Vietnams, higher in Jun and Jul months. Peak in jul. Sept to Dec relatively constant, Jan to May, gradual increase. Jun higher jump. Arrival count peak at jul
```{r}
#| fig-height: 6
tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  gg_subseries(Arrivals)
```



### ACF plots
how the time series correlated.
only one variable - which is your time series variable

auto correlation plots
create 2 time lags...
find the correlation between them
- then you have multiple variables.
- get the second variable by shifting lag=1
note the graph below starts from lag=1 (t vs t-1)


China - 6 mths period.
Italy - 12 peak. weak correlation at 1st mth ~0.3
Note!!! 95% confidence level - should go above the blue line.
China case, vietnam case, most/all the lags is statistically significant.
China and vietnam, correlation decreases then increases again. The period different. Vietnam - 12 mths. China - 6 mths period
UK lag t-1 significant... the rest of the year not sigificant, then 12 mths significant again
- which means trend is not significant at all
- drop down very fast
- seasonal not important. (for UK and Italy)


```{r}
#| fig-height: 6
tsibble_longer %>%
  filter(`Country` == "Vietnam" |
         `Country` == "Italy" |
         `Country` == "United Kingdom" |
         `Country` == "China") %>%
  ACF(Arrivals) %>%
  autoplot()
```

