---
title: "In-class Exercise 4"
author: "Sindy"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Getting started

### Installing and Loading R packages

```{r}
pacman::p_load(haven, SmartEDA, tidyverse, tidymodels, ggridges)
```

```{r}
exam_data <- read_csv("./data/Exam_data.csv")
```

## Box plot

```{r}
ggplot(data = exam_data,
       aes(x = ENGLISH,
           y = CLASS)) + 
  geom_boxplot()
```

## Ridgeline and Boxplot

Boxplots provide essential summary statistics—medians, quartiles, and whiskers—but they are incomplete, they don’t reveal the full distribution of the data. 

In contrast, ridgeline plots illustrate the underlying density, highlighting features like multiple peaks (as seen in classes 3G and 3H) and outliers (evident in class 3F). 

Combining both provides a more complete view by showing both the summary metrics and the detailed distribution.

```{r}
# Create the combined plot
ggplot(data = exam_data, aes(x = ENGLISH, y = CLASS)) +
  geom_density_ridges(
    scale = 1.1,        # Adjusts the vertical scaling of the ridgelines
    alpha = 0.5,    
    fill = "lightblue"
  ) +
  geom_boxplot(
    width = 0.15,
    position = position_nudge(y = -0.2),  
    outlier.colour = "red",              
    alpha = 0.7                       
  ) +
  theme_minimal() +
  labs(
    title = "Combined Plot of English Scores by Class",
    x = "English Score",
    y = "Class"
  )
```
